hatchR: A toolset to predict hatch and emergence phenology in wild fishes

### 2025-02-28 inst/manuscript/paper.Rmd
introduction edits

---
title: 'hatchR: A toolset to predict hatch and emergence phenology in wild fishes'
tags:
  - R
  - fishes
  - developmental phenology
authors:
  - name: Morgan M. Sparks
    orcid: 0000-0003-0787-2218
    affiliation: 1 # (Multiple affiliations must be quoted)
  - name: Eli Felts
    orcid: 0000-0002-5888-9266
    affiliation: 2
  - name: Bryan M. Maitland
    orcid: 0000-0002-4491-5064
    affiliation: 1
affiliations:
 - name: US Forest Service, Rocky Mountain Research Station, Boise, ID, USA
   index: 1
 - name: US Fish and Wildlife Service, Idaho Fish and Wildlife Conservation Office, Orofino, ID, USA
   index: 2
citation_author: Sparks et al.
date: 07 July 2025
year: 2025
output: rticles::joss_article
csl: apa.csl
journal: JOSS
editor_options:
  chunk_output_type: console
bibliography: ms_refs.bib
---

```{r, include=FALSE}
library(tidyverse)
library(hatchR)
library(ggridges)
```

@@

Understanding the timing of key life history events is necessary for managing and conserving populations. Historically, models to predict hatch and emergence timing for fishes were difficult to employ in wild settings because average incubation temperature was needed as the primary parameter in predictive models. However, recent improvements to these techniques reworked models such that they could be applied in wild environments as long as users had data for when adult fish spawned and a record of average daily temperature over the course of development. Despite these improvements, their application remains limited due to few parameterizations for varying species, being largely limited to salmonids. Here we present **hatchR**, a software ecosystem that allows users to predict hatch and emergence timing for wild fishes, as well as additional tools to aid in those analyses. **hatchR** allows users to leverage popular historic parameterizations for phenological models or to easily implement custom parameterizations using data not included in the package. **hatchR** is also distributed in two forms—an open source R package for maximum customization, as well as an HTML graphical-user-interface web application for individuals not familiar with scripting languages. To demonstrate potential uses, we present two case studies as likely applications for this software. **hatchR** promises to open many exciting avenues in research and management of fishes during their early life history.

@@

As primarily poikilothermic organisms, the development and growth of fishes is tightly linked with the temperature of their ambient environment. This close relationship has allowed researchers to generate statistical models that allow the prediction of developmental phenology with high accuracy and precision. These models were typically developed in aquaculture settings and their initial formulations were not applicable to wild populations because they assumed a constant temperature over the course of development [@beacham1990; @mcphail1979; @alderdice1978]. However, @sparks2019 reformulated this approach as an "Effective Value model", in which the input was daily average temperature after a parent spawned and fish would either hatch or emerge when effective values cumulatively summed to one.

The resulting effective value approach has now been widely applied in Salmonids for which parameterizations from aquaculture were readily available—for example Pacific Salmon (*Oncorhynchus spp.*) models developed by @beacham1990 have been applied to various species and populations [@adelfio2024; @adelfio2019; @kaylor2021] while models developed for Bull Trout (*Salvelinus confluentus*) by @mcphail1979 were extended by @austin2019. Despite growing popularity, applications have been largely limited within Salmonids, presumably because parameterizations for such models already existed due to their wide use in aquaculture and their general popularity as sport and commercial fish.

To bridge the gap between the application of one-off effective value model applications within individual studies and the lack of parameterization for other species, we developed the software ecosystem, **hatchR**. Specifically, **hatchR** allows users to input standard raw or summarized temperature datasets that are commonly collected in wild settings, run basic checks on those data, use built-in parameterizations like those from @beacham1990 or @sparks2017, develop custom models from their own or published temperature and phenological data, and predict hatch and emergence timing using these models in the effective value framework.

To widen the user application of these methods, we distribute two user-interfaces for **hatchR**. The first is a R package (@r:ther) distributed via CRAN that allows users the most customizable application for these methods. The R package is especially powerful as it allows users to automate their analyses over multiple variables such as phenology type, multiple spawn dates, or different habitats with varying thermal regimes. These variable approaches are outlined in the package documentation on **hatchR**'s website. Alternatively, we also distribute a Shiny application (@chang2024) in the form of an HTML-based web tool to interact with many of **hatchR**'s functions in a graphical-user-interface. The Shiny form trades-off some of automative power for user simplicity, while still allowing users to leverage much of the functionality of **hatchR**'s R package. Below, we present the basic overview of the software and multiple case studies of how it may be applied.

```{r echo=FALSE, message=FALSE}
### 
library(tidyverse); library(hatchR); library(ggdist)
```

@@

**hatchR** is meant to primarily be a tool for predicting phenology. In this sense, we limit functionality to these applications and provide minimal data checking and plotting help. This decision is in part driven by the diversity of data types that users may import and the difficulty in addressing all those data types with respect to various data checks. In other words, we expect users to know their data better than we do and to check it accordingly. We do provide two basic data check functions discussed in the Checking Data section. Similarly, we provide limited functionality for plotting results, but provide examples of how to build custom visualizations from output, specifically in R. For the Shiny application, we provide a base output plot, but the ability to download your results for custom plotting in programs of the user's choice. Finally, we provide brief summaries of general applications of **hatchR** below, but encourage users to visit articles hosted on the software webpage that extensively outline primary functions and applications, especially automating the application of predicting phenology across multiple variables.

```{r echo = FALSE, out.height = '50%', fig.align = 'center', fig.cap= "Full workflow for hatchR. Data processes are demonstrated with beige circles and pakcage functions with white rectangles."}
knitr::include_graphics("flowchart.png")
```

@@

Water temperature datasets collected for wild environments are often either 1.) already summarized by day (*i.e.*, mean daily temperature) or, 2.) in a raw format from something like a HOBO TidbiT where readings are taken multiple times per day, which can be summarized into a mean daily temperatures. Alternatively, new statistical models like that of @siegel2023 could be similarly implemented.

Fundamentally, **hatchR** assumes you have input data with two columns: a date column, giving the date (and often time) of a temperature measurement, and a temperature column, giving the associated temperature measurement (in centigrade). Other columns are okay to include, but these two columns (with any column name—just *without* spaces) are required. We expect your data to look something like this:

| date       | temperature |
|------------|-------------|
| 2000-01-01 | 2.51        |
| ...        | ...         |
| 2000-07-01 | 16.32       |
| ...        | ...         |
| 2000-12-31 | 3.13        |

**hatchR** assumes you've checked for missing records or errors in your data as it *will function with gaps*, so it's important to go through the data checks discussed below, as well as your own validity checks. **hatchR** can use values down to freezing (e.g, 0 °C), which returns extremely small effective values, and time to hatch or emerge may be \> 1 year. In these cases, we suggest users consider how much of that data type is reasonable with their data.

@@

We demonstrate the utility of the summary and check functions `summarize_temp()`, `plot_check_temp()`, and `check_continuous()` using a simulated year-long dataset with temperature readings every thirty minutes.

```{r}
# create date object for a year with 30 min reading intervals
dates <- seq(from = ymd_hms("2000-07-01 00:00:00"),
             to = ymd_hms("2001-06-30 23:59:59"), length.out = 17568)

@@

dim(year_sim)

@@

First, we recommend checking imported data for any outliers or strange inputs using `plot_check_temp()`

```{r, fig.cap= "Full output of the plot_check_temp() visual data check function from hatchR. Note that the red and blue lines are customizable by the user to create outlier thresholds for their visual check."}
plot_check_temp(data = year_sim,
                dates = date,
                temperature = temp,
                temp_min = 0, # temp_min and max lines are 
                              # user customizable
                temp_max = 25)
```

@@

Users can either use Salmonid models from `model_table` included in the package and the Shiny app. As discussed, these models are included because parameterizations already existed in the literature. However, these parameterizations are limited to Pacific Salmon and Bull Trout. To widen the application of the effective value approach, we include a `fit_model()` function and is species agnostic (as long as develop generally follows the power law.

The function `fit_model()` uses data in which average incubation temperature (°C) and days to phenological event (as two vectors) are the inputs and estimates parameter coeficients for *log~e~a* and *b* using `stats::nls()`

@@

#### Fitting models for other fishes

Below, we demonstrate how the `fit_model()` function may be used to create custom parameterizations for species beyond the Salmonids in the `model_table` included in the package. We include parameterizations from three warm-water species to demonstrate the `fit_model()` utility for fishes beyond the scope of the original effective value approach. These parameterizations are for commonly cultured sportfishes including Smallmouth Bass (*Micropterus dolomieu*), Channel Catfish (*Ictalurus punctatus*) from @small2001, and Lake Sturgeon (*Acipenser fulvescens*) from @smith2005. We demonstrate parameterization for just Smallmouth Bass here for concision, but the full code set for all species is available at <https://github.com/bmait101/hatchR/blob/master/inst/manuscript/paper.Rmd>.

@@

```{r}
###  make temp regime
set.seed(123)
# create random temps and corresponding dates
temps_sim <- sort(rnorm(n =30, mean = 16, sd = 1), decreasing = FALSE)
dates_sim <-  seq(from = ymd("2000-07-01"),
             to = ymd("2000-07-31"), length.out = 30)

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap= "Days to hatch as predicted using custom parameterizations for Channel Catfish (blue), Lake Sturgeon (black, and Smallmouth Bass (green) as generated by the fit_model() function in hathcR. Models are all run using a spawn date starting June 1 for a randomly generated ascending thermal regime with mean temperature 16 °C"}
### predict_phenology

@@

To illustrate model selection and phenology prediction we will recreate a small portion of the analysis done by @sparks2019 using the `woody_island` dataset included in this package . We will predict both hatch and emergence timing, so we will obtain a model expression for each using `model_select()`.

```{r}
sockeye_hatch_mod <- model_select(
  author = "Beacham and Murray 1990", 
  species = "sockeye", 
  model = 2, 
  development_type = "hatch"
  )
```

@@

```{r, warning=FALSE}
WI_hatch <- predict_phenology(
  data = woody_island,
  dates = date,
  temperature = temp_c,
  spawn.date = "1990-08-18",
  model = sockeye_hatch_mod
  )
```

And then look inside the returned object to see days to hatch and development period:

```{r}
WI_hatch$days2done
WI_hatch$dev.period
```

#### Understanding your results

@@

```{r, eval=FALSE}
str(WI_hatch)
```

`WI_hatch$days2done` outputs the predicted days to develop.

`WI_hatch$dev.period` is a 1x2 dataframe with the dates corresponding to when your fish's parent spawned (which you input with `predict_phenology(spawn.date = ...)`) and the date when the fish is predicted todevelop.

`WI_hatch$ef.vals` is a vector of each day's effective value as evaluated using whatever model is chosen.

`WI_hatch$ef.tibble` is a *n* x 4 tibble (*n* = number of days to hatch or emerge) and the columns are the date, each day's temperature and effective value, and the cumulative sum of the effective values. The `ef.tibble` object is meant to serve as the basis for users to make custom figures for their data beyond the functionality we discuss below.

#### Plotting phenology

@@

```{r, fig.cap= "Output of the plot_phenology() function from hatchR. The function plots the output of the predict_phenology() function and displays time to develop, as well as the ambient thermal regime (blue), scaled daily effective value (green), and the scaled cumulative sum of effective values (orange)."}
plot_phenology(WI_hatch)
```

@@

A common management scenario where developmental phenology might be useful would be trying to understand if fish might be free-moving before some management action. For instance, will have fish have emerged from redds when a stream section has been opened to grazing or bridge decommissioning will commence?

In this scenario, we will consider the grazing example and Bull Trout, a threatened fish in the United States under the Endangered Species Act [@nolfi2024], and the East Fork Salmon River, a key Bull Trout population in the upper Salmon River watershed. The fisheries manager there wants to know if fish will likely be out of the gravel and free-swimming by June 1st. In this system, it is expected that Bull Trout will be done spawning by the end of September, so we'll consider the last possible spawn date as September 30th.

We demonstrate this first case study using the graphical user interface portion of the **hatchR** ecosystem found at <https://elifelts.shinyapps.io/hatchR/>. Users will first upload their data with the `Import Data` window, which requires them to select their file on their personal computer, provide the program with the columns corresponding for dates and temperatures, and then provide the format in which dates are coded (*e.g.*, year-month-day or day-month-year). Once data is uploaded the program automatically plots the user's data using `plot_check_temp()` in the background and provides them the outputted graphical check. After uploading and checking data, the user switches to the `Model Phenology` window. In this circumstance, we use the preloaded parameterization for bull trout from @austin2019 with the `Existing` button for model selection, which the user selects with the various drop down options in the menu. After the model is selected, the user can choose multiple spawn dates from the interactive calendar provided. We show results for spawning for September 30th as indicated in the example above. Once dates are chosen, a table entry for each spawn date is outputted in the `Phenology Summaries` tab and corresponding plot with data from each spawn date in the `Timeline Plot` tab. Output from predicting phenology and the resulting figure are downloadable from their respective tabs. The process is demonstrated in full in Figure XXX, but the interface is described more completely in the Articles on **hatchR**'s website <https://bmait101.github.io/hatchR/>.

@@

# bull trout hatch model
bt_hatch <- model_select(
  development_type = "hatch",
  author = "Austin et al. 2017",
  species = "bull trout",
  model = "MM"
)
```

We need to setup our models and data for this analysis, which we don't show those steps here for the sake of concision in this manuscript, however they are demonstrated in `paper.Rmd` included in the GitHub repository for **hatchR**. After the setup, we can easily map `predict_phenology()` across all putative spawning sites and three spawn dates (September 1-Early Spawning, September 15-Peak Spawning, and September 31-Late Spawning), the results of which are presented in Figure XXX.

```{r}
hatch_res <- isaak_summ_bt |> 
  mutate(
    dev_period = map2(
      summ_obj, spawn_dates, # map across our site object and spawn dates
      predict_phenology,
      temperature = daily_temp,
      model = bt_hatch,
      dates = date
      ) |> 
      map_df("dev.period") |> # pull out just dev.period results
      list()
    ) |> 
  select(site, dev_period) |> # just select the columns we want
  unnest(cols = c(dev_period)) |> # unnest everything
  mutate(days_to_hatch = stop - start) # make a new column of days to hatch
```

```{r, message=FALSE, echo=FALSE, fig.cap= "Densities of predicted days to hatch for Bull Trout for 139 putative spawning locations across Idaho for four years of shared temperature data. Early spawning considers if a parent spawned September 1, Peak-September 15, and Late-September 30."}
cut_ints <- date(
  c(
    "2011-08-31", "2012-08-31",
    "2013-08-31", "2014-08-31",
    "2015-08-31"
  )
)

@@

We thank Laura Koller for her help designing the **hatchR** logo. Dan Isaak provided useful discussion about model development and temperature datasets.

@@

>>>>

---
title: ""
output: pdf_document
header-includes:
   - \usepackage{lineno}
   - \linenumbers
bibliography: ms_refs.bib
csl: apa.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## hatchR: A toolset to predict when fish hatch and emerge

Morgan M. Sparks^1,4^, Eli A. Felts^2^, Allison G. Swartz^3^, Bryan M. Maitland^1^

^1^Rocky Mountain Research Station, U.S. Forest Service, Boise, ID, USA

^2^Idaho Fish and Wildlife Conservation Office, U.S. Fish and Wildlife Service, Orofino, ID, USA

^3^College of Forestry, Oregon State University, Corvallis, OR, USA

^4^Corresponding Author: Morgan M. Sparks, Morgan.Sparks\@usda.gov

keywords: developmental phenology, effective value model, R, Shiny

\newpage

@@

Understanding the timing of key life history events is essential for effective fish conservation and management. Traditionally, predicting hatch and emergence timing in wild fish populations was challenging due to the reliance on average incubation temperature as a primary model parameter, which is often difficult in to obtain in natural settings. Recent advancements have refined these models, enabling their application in wild environments using spawning dates and daily water temperature records. However, their broader use remains constrained by a lack of parameterizations for many species, with most applications focused on salmonids. Here we introduce **hatchR**, a software ecosystem designed to predict hatch and emergence for a wide range of wild fishes. **hatchR** offers users access to established phenological models and the flexibility to incorporate custom parameterizations using external datasets. The software is available in two formats: an open-source R package for advanced customization and an HTML-based graphical user interface for those unfamiliar with scripting. To illustrate its utility, we present two case studies demonstrating its application in research and management. By expanding assess to predictive modeling tools, **hatchR** has the potential to advance studies of fish early life history and support conservation efforts across diverse species.

\newpage

@@

As poikilothermic organisms, fish development and growth are closely tied to ambient environment. This strong relationship has enabled researchers to generate statistical models that predict developmental phenology with high accuracy. Historically, these models were formulated in aquaculture settings under the assumption of constant temperature throughout development [@beacham1990; @mcphail1979; @alderdice1978], limiting their applicability to wild populations. However, @sparks2019 reformulated this approach into an "Effective Value model", which instead uses daily average temperature after spawning, predicting hatch or emergence when cumulative effective values reach a threshold of one.

This effective value approach has since been widely applied to salmonids, for which aquaculture-derived parameterizations were readily available. For example, Pacific Salmon (*Oncorhynchus* spp.) models developed by @beacham1990 have been applied across various species and populations [@adelfio2024; @adelfio2019; @kaylor2021], while Bull Trout (*Salvelinus confluentus*) models from @mcphail1979 were extended by @austin2019. Despite its growing adoption, applications of the effective value model remian largely confined to salmonids, likelu due to the availability of existing parameterizations and the commercial and recreational importance of these species.

To extrend these modeling capabilities beyond salminds and facilitate broader applications, we developed **hatchR**, a software ecosystem designed to predict hatch and emergence timing for wild fish populations. **hatchR** enables users to input standard raw or summarized water temperature datasets commonly collected in field settings, conduct basic data validation, and apply built-in parameterizations such as those from @beacham1990 or @sparks2017. Users can also develop custom models using their own or published temperature and phenological data within the effective value framework.

To maximize accessibility, **hatchR** is available in two formats. The first is a R package, **hatchR**, distributed via CRAN (REF), providing advanced customization and automation for analyzing mutiple variables, such as phenology type, spawn timing, or thermal regimes. Comprehensize documentation is available on the **hatchR** website (<https://bmait101.github.io/hatchR/>). The second is a Shiny-based web application (@chang2024), offering a graphical user interface for those unfamiliar with R, balancing ease of use with much of the R package's core functionality. Below, we provide and overview of **hatchR** and present case studies demonstrating its application in research and management.

```{r echo=FALSE, message=FALSE}
### 
library(tidyverse); library(hatchR); library(ggdist); library(ggridges)
```

@@

**hatchR** is meant to primarily be a tool for predicting phenology. In this sense, we limit functionality to these applications and provide minimal data checking and plotting help. This decision is in part driven by the diversity of data types that users may import and the difficulty in addressing all those data types with respect to various data checks. In other words, we expect users to know their data better than we do and to check it accordingly. We do provide two basic data check functions discussed in the Checking Data section. Similarly, we provide limited functionality for plotting results, but provide examples of how to build custom visualizations from output, specifically in R. For the Shiny application, we provide a base output plot, but the ability to download your results for custom plotting in software of the user's choice. Finally, we provide brief summaries of general applications of **hatchR** below (Figure 1), but encourage users to visit articles hosted on the software webpage that extensively outline primary functions and applications, especially automating the application of predicting phenology across multiple variables.

```{r, echo = FALSE, out.height = '75%', fig.align = 'center', fig.cap="hatchR workflow. Data inputs are represented by folders, data processes by filled circles, hatchR functions as plain text rectangles, and decisions choices as italicized text."}
knitr::include_graphics("./workflow.png")
```

@@

Water temperature data sets collected for wild environments are often either 1.) already summarized by day (*i.e.*, mean daily temperature) or, 2.) in a raw format from something like a HOBO TidbiT where readings are taken multiple times per day, which can be summarized into a mean daily temperatures. Alternatively, new statistical models like that of @siegel2023 could be similarly implemented.

Fundamentally, **hatchR** assumes you have input data with two columns: a date column, giving the date (and often time) of a temperature measurement, and a temperature column, giving the associated temperature measurement (in centigrade). Other columns are okay to include, but these two columns (with any column name—just *without* spaces) are required. Data is expected to be formatted similar to Table 1.

```{r kabletable, echo=FALSE }
df = data.frame(date = c("2000-01-01",
                         "...",
                         "2000-07-01",
                         "...",
                         "2000-12-31"),
                temperature= c(2.51,
                        "...",
                        16.32,
                        "...",
                        3.13
                        ))

# kable can alse be used for creating tables
knitr::kable(df,
             booktabs=TRUE,
             caption = "Example temperature data for use in hatchR.")
```

**hatchR** assumes you've checked for missing records or errors in your data as it *will function with gaps*, so it's important to go through the data checks discussed below, as well as your own validity checks. **hatchR** can use values down to freezing (e.g, 0 °C), which returns extremely small effective values, and time to hatch or emerge could extend to a year or more. In these cases, we suggest users consider how much of that data type is reasonable with their data.

@@

We demonstrate the utility of the summary and check functions `summarize_temp()`, `plot_check_temp()`, and `check_continuous()` using a simulated year-long data set (`year_sim`, the dimensions of the data and first 6 rows are shown) with temperature readings every thirty minutes.

```{r, echo = FALSE}
# create date object for a year with 30 min reading intervals
dates <- seq(from = ymd_hms("2000-07-01 00:00:00"),
             to = ymd_hms("2001-06-30 23:59:59"), length.out = 17568)

@@

@@

```{r}
#year_sim data dimensions
dim(year_sim)
#fist 6 rows of year_sim
head(year_sim)
```

First, we recommend checking imported data for any outliers or strange inputs using `plot_check_temp()` (Figure 2).

```{r, echo = TRUE, fig.cap="Output of hatchR function `plot_check_temp()`, which is used as a visual data check on the raw `year_sim` data set. Users can set custom thresholds for minimum and maximum temperatures (dashed lines)."}
plot_check_temp(data = year_sim,
                dates = date,
                temperature = temp,
                temp_min = 0, # temp_min and max lines are 
                              # user customizable
                temp_max = 25)
```

@@

Users can either use Salmonid models from `model_table` or generate custom models suing `fit_model()` in both R and Shiny deployments of **hatchR**. As discussed, the models in `model_table` are included because parameterizations already existed in the literature, however, these parameterizations are limited to Pacific Salmon and Bull Trout (though see @quinn2018 pg. 183 for numerous other Salmonids). Additionally, we limit available models in `model_table` because they are well vetted models, with experimental ranges spanning 2-17 °C. Custom models we demonstrate later with `fit_model()` have much narrower parameterization ranges and, as such, we choose to keep those models out of `model_table` so as to require users to consider whether the custom models they are parameterizing fit the data over which they are predicting.

To widen the application of the effective value approach, we include a `fit_model()` function, which is species agnostic (as long as development generally follows the power law). The function `fit_model()` uses data in which average incubation temperature (°C) and days to phenological event (as two vectors) are the inputs. The function estimates parameter coefficients for *log~e~a* and *b* using `stats::nls()`

@@

### Fitting models for other fishes

We demonstrate how the `fit_model()` function may be used to create custom parameterizations for species beyond the Salmonids in the `model_table` included in the package. We include parameterizations from three warm-water species to demonstrate the `fit_model()` utility for fishes beyond the scope of the original effective value approach. These parameterizations are for commonly cultured sportfishes including Smallmouth Bass (*Micropterus dolomieu*) [@webster1948], Channel Catfish (*Ictalurus punctatus*) [@small2001], and Lake Sturgeon (*Acipenser fulvescens*) [@smith2005]. We show parameterization for just Smallmouth Bass here for concision, but the full code set for all species is available at in the [paper.Rmd](https://github.com/bmait101/hatchR/blob/master/inst/manuscript/paper_pdf/paper.Rmd) in the project GitHub.

@@

```{r, echo=FALSE}
###  make temp regime
set.seed(123)
# create random temps and corresponding dates
temps_sim <- sort(rnorm(n =30, mean = 16, sd = 1), decreasing = FALSE)
dates_sim <-  seq(from = ymd("2000-07-01"),
             to = ymd("2000-07-31"), length.out = 30)

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Predicted days to hatch for three warmwater species with custom parameterizations using a random thermal regime with an ascending thermograph with a mean temperature of 16 °C over 30 days."}
### predict_phenology

@@

To illustrate model selection and phenology prediction we will recreate a small portion of the analysis done by @sparks2019 using the `woody_island` data set included in this package. We will predict both hatch and emergence timing, so we will obtain a model expression for each using `model_select()`, which calls built-in models from `model_table`.

```{r, echo = TRUE}
sockeye_hatch_mod <- model_select(
  author = "Beacham and Murray 1990", 
  species = "sockeye", 
  model = 2, 
  development_type = "hatch"
  )
```

@@

```{r, warning=FALSE, echo = TRUE}
WI_hatch <- predict_phenology(
  data = woody_island,
  dates = date,
  temperature = temp_c,
  spawn.date = "1990-08-18",
  model = sockeye_hatch_mod
  )
```

And then look at the returned object to see days to hatch and development period:

```{r, echo = TRUE}
WI_hatch$days_to_develop
WI_hatch$dev.period
```

### Understanding your results

@@

```{r, eval=FALSE, echo = TRUE}
str(WI_hatch)
```

`WI_hatch$days_to_develop` outputs the predicted days to develop.

`WI_hatch$dev.period` is a 1x2 dataframe with the dates corresponding to when your fish's parent spawned (which you input with `predict_phenology(spawn.date = ...)`) and the date when the fish is predicted to develop.

`WI_hatch$ef_table` is a *n* x 5 tibble (*n* = number of days to hatch or emerge) and the columns are a row index, the date, each day's temperature and effective value, and the cumulative sum of the effective values. The `ef_table` object is meant to serve as the basis for users to make custom figures for their data beyond the functionality we discuss below.

`WI_hatch$model_specs` provides details about the model used for predicting phenology, which varies depending on if the model was generated from `model_select()` or `fit_model()`, but most importantly contains the model expression (*e.g.*, the formula) used for prediction.

### Plotting phenology

@@

```{r, fig.cap="Output of `plot_phenolgy()` function using predicted hatch time from `woody_island` data set."}
plot_phenology(WI_hatch)
```

@@

A common management scenario where developmental phenology might be useful would be trying to understand if fish might be free-moving before some management action. For instance, will fish have emerged from redds when a stream section has been opened to grazing or road work?

In this scenario, we will consider the road work example and Bull Trout, a threatened fish in the United States under the Endangered Species Act [@nolfi2024], and the Crooked River, a key Bull Trout population in the Boise River watershed. In this hypothetical scenario, the Forest Fisheries Biologist wants to know if fish will likely be out of the gravel and free-swimming by June 1st as Bull Trout are particularly sensitive to sediment. In this system, it is expected that Bull Trout will be done spawning by the end of September, so we'll consider the last possible spawn date as September 30th.

We demonstrate this first case study using the graphical user interface portion of the **hatchR** ecosystem found at [https://elifelts.shinyapps.io/hatchR_demo/](https://elifelts.shinyapps.io/hatchR/){.uri}. Users will first upload their data with the `Import Data` window, which requires them to select their file on their personal computer, provide the program with the columns corresponding for dates and temperatures, and then provide the format in which dates are coded (*e.g.*, year-month-day or day-month-year). Data used in this case study are from the `crooked_river` data set. It is included as a demo data set in Shiny, but can also be written out from the package or accessed at <https://github.com/bmait101/hatchR/blob/master/extdata/crooked_river.csv/>. Once data is uploaded the program automatically plots the user's data using `plot_check_temp()` in the background and provides them the outputted graphical check. After uploading and checking data, the user switches to the `Model Phenology` window. In this circumstance, we use the preloaded parameterization for bull trout from @austin2019 with the `Existing` button for model selection, which the user selects with the various drop down options in the menu. After the model is selected, the user can choose multiple spawn dates from the interactive calendar provided. We show results for spawning for September 30th as indicated in the example above. Once dates are chosen, a table entry for each spawn date is outputted in the `Phenology Summaries` tab and corresponding plot with data from each spawn date in the `Timeline Plot` tab. Output from predicting phenology and the resulting figure are downloadable from their respective tabs. The process is demonstrated in the included ***supplementary video file???***, but the interface is described more completely on **hatchR**'s Shiny website.

@@

# bull trout hatch model
bt_hatch <- model_select(
  development_type = "hatch",
  author = "Austin et al. 2019",
  species = "bull trout",
  model = "MM"
)
```

We need to setup our models and data for this analysis, which we don't demonstrate here for the sake of concision, however they are demonstrated in `paper.Rmd` included in the GitHub repository for **hatchR**. After the setup, we can easily map `predict_phenology()` across all putative spawning sites and three spawn dates (September 1-Early Spawning, September 15-Peak Spawning, and September 31-Late Spawning), the results of which are presented in Figure 5.

```{r, echo = TRUE}
hatch_res <- isaak_summ_bt |> 
  mutate(
    dev_period = map2(
      summ_obj, spawn_dates, # map across our site object and spawn dates
      predict_phenology,
      temperature = daily_temp,
      model = bt_hatch,
      dates = date
      ) |> 
      map_df("dev.period") |> # pull out just dev.period results
      list()
    ) |> 
  select(site, dev_period) |> # just select the columns we want
  unnest(cols = c(dev_period)) |> # unnest everything
  mutate(days_to_hatch = stop - start) # make a new column of days to hatch
```

```{r, message=FALSE, echo=FALSE,  out.width='75%', fig.align='center', fig.cap="Predicted days to hatch for 139 putative bull trout populations over three spawning periods (Early = September 1, Peak = September 15, Late = September 30) and four years of temperature data."}
cut_ints <- date(
  c(
    "2011-08-31", "2012-08-31",
    "2013-08-31", "2014-08-31",
    "2015-08-31"
  )
)

@@

We thank Laura Koller for her help designing the **hatchR** logo. Dan Isaak provided useful discussion about model development and temperature data sets.

@@

# Conflicts of Interest

The authors declare no known conflicts of interest.

# Data Availability

**hathcR** is fully open source and reproducible. Source code and data can be found at <https://github.com/bmait101/hatchR/>. The latest version will be archived upon acceptance of the manuscript.

# Ethics Statement

All data was derived from pre-published sources or created synthetically.

# Funding

The authors declare no funding sources.
